---
title: "Ulcerative Colitis Metabolomics sPLSDA"
author: "Isabel Prado-Tucker"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Data source: (The Inflammatory Bowel Disease Multi'omics Database)[https://ibdmdb.org/]

### Data selection: n=120 (60 UC and 60 nonIBD samples selected with M/F split). Only C8-pos method data selected from metabolomic dataset.

### Data preparation: Selected data was run through xmsPanda regression analysis to determine selected features. This analysis is performed on the regression selected features data.

```{r, include=FALSE}
library(data.table)
library(mixOmics)
set.seed(5249) # for reproduction
```

```{r, include=FALSE}
# Prepare and import data
metab_sf <- fread("~/Desktop/stanford/usefulData/100n/UC_120_lmregresults_selectedfeatures.csv")
samples <- fread("~/Desktop/stanford/usefulData/100n/sample_metadata.csv") # Selected samples
diag <- as.factor(samples$diagnosis)
```

```{r, include=FALSE}
# Flip metabo data 
flipped_metabo <- as.data.frame(t(metab_sf))

# Only select matrix
flipped_metabo <- flipped_metabo[7:nrow(flipped_metabo),]

# Convert to matrix
metab_matrix <- as.matrix(flipped_metabo) # Stop at 1:71
metab <- matrix(as.numeric(metab_matrix), ncol = ncol(metab_matrix))
metab <- setNames(metab, colnames(metab_matrix))
colnames(metab) <- colnames(metab_matrix)
colnames(metab) <- sub("V", "X", colnames(metab))

dim(metab)
```

```{r, include=FALSE}
library(dplyr)
set.seed(5249)
train <- sample(c(TRUE, FALSE), nrow(metab), replace=TRUE, prob=c(0.6,0.4))

# store matrices into training and test set:
metab.train <- metab[train, ]
metab.test <- metab[!train,]
diag.train <- diag[train]
diag.test <- diag[!train]
```

***

## Initial sPLS-DA model

#### FIGURE 1: Barplot of the variance each principal component explains of the UC metabolomic data.

```{r echo=FALSE}
pca.metabo = pca(metab, ncomp = 10, center = TRUE, scale = TRUE) # run pca method on data
plot(pca.metabo)  # barplot of the eigenvalues (explained variance per component)
```
```{r}
plotLoadings(pca.metabo) # plot the bar plot for the first principal component

# plot the bar plot for the second principal component
plotLoadings(pca.metabo, comp = 2) 
```

#### FIGURE 2: Preliminary (unsupervised) analysis with PCA on the UC metabolomic data

```{r echo=FALSE}
plotIndiv(pca.metabo, group = diag, ind.names = FALSE, # plot the samples projected
  legend = TRUE, title = 'PCA on Metabolomics, comp 1 - 2') # onto the PCA subspace
```

```{r, include=FALSE}
metabo.splsda <- splsda(metab, diag, ncomp = 10)  # set ncomp to 10 for performance assessment later
```

```{r}
# plot the bar plot using the highest value of the median to colour the bars
plotLoadings(metabo.splsda, contrib = 'max', method = 'median') 
```

```{r, include=FALSE}
# use the max.dist measure to form decision boundaries between classes based on PLS-DA data
background = background.predict(metabo.splsda, comp.predicted=2, dist = "max.dist")
```

#### FIGURE 3: Sample plots of the UC metabolomic data after a basic PLS-DA model was operated on this data.

```{r, echo=FALSE}
# plot the samples projected onto the first two components of the PLS-DA subspace
plotIndiv(metabo.splsda , comp = 1:2, 
    group = diag, ind.names = FALSE,  # colour points by class
    ellipse = TRUE, # include 95% confidence ellipse for each class
    legend = TRUE, title = '(a) PLSDA with confidence ellipses')

# plot the samples projected onto the first two components of the PLS-DA subspace
plotIndiv(metabo.splsda, comp = 1:2,
          group = diag, ind.names = FALSE, # colour points by class
          background = background, # include prediction background for each class
          legend = TRUE, title = " (b) PLSDA with prediction background")
```

***

### Tuning sPLS-DA

```{r, echo=FALSE}
# undergo performance evaluation in order to tune the number of components to use
perf.splsda.metabo <- perf(metabo.splsda, validation = "Mfold", 
                          folds = 10, nrepeat = 50, # use repeated cross-validation # increase folds and repeat for accuracy
                          progressBar = FALSE, auc = TRUE) # include AUC values

# plot the outcome of performance evaluation across all ten components
plot(perf.splsda.metabo, col = color.mixo(5:7), sd = TRUE,
     legend.position = "horizontal")
```

```{r, include=FALSE}
perf.splsda.metabo$choice.ncomp # what is the optimal value of components according to perf()
```

#### FIGURE 4: Tuning the number of components in PLS-DA on the UC metabolomic data

```{r}
# grid of possible keepX values that will be tested for each component
list.keepX <- c(1:10,  seq(20, 300, 10))

# undergo the tuning process to determine the optimal number of variables
tune.splsda.metabo <- tune.splsda(metab, diag, ncomp = 4, # calculate for first 4 components 
                                 validation = 'Mfold',
                                 folds = 10, nrepeat = 50, # use repeated cross-validation
                                 dist = 'max.dist', # use max.dist measure
                                 measure = "BER", # use balanced error rate of dist measure
                                 test.keepX = list.keepX,
                                 cpus = 2) # allow for paralleliation to decrease runtime

plot(tune.splsda.metabo, col = color.jet(4)) # plot output of variable number tuning # Had to change to 3 to reflect ncomp value
```

```{r, include=FALSE}
tune.splsda.metabo$choice.ncomp$ncomp # what is the optimal value of components according to tune.splsda()
```

```{r, include=FALSE}
tune.splsda.metabo$choice.keepX # what are the optimal values of variables according to tune.splsda()
```

```{r, include=FALSE}
optimal.ncomp <- tune.splsda.metabo$choice.ncomp$ncomp
optimal.keepX <- tune.splsda.metabo$choice.keepX[1:optimal.ncomp] # Originally [1:optimal.ncomp]
```

***

## Creating the final model

```{r, include=FALSE}
# form final model with optimised values for component and variable count
final.splsda <- splsda(metab, diag, 
                       ncomp = optimal.ncomp, #suppesed to be optimal.ncomp (1)
                       keepX = optimal.keepX)
```

```{r, echo=FALSE}
plotIndiv(final.splsda, comp = c(1,1), # plot samples from final model, c(1,2)
          group = diag, ind.names = FALSE, # colour by class label
          ellipse = TRUE, legend = TRUE, # include 95% confidence ellipse
          title = 'sPLS-DA on UC, comp 1')

plotIndiv(final.splsda, comp = c(1,2), # plot samples from final model, c(1,2)
          group = diag, ind.names = FALSE, # colour by class label
          ellipse = TRUE, legend = TRUE, # include 95% confidence ellipse
          title = 'sPLS-DA on UC, comp 1 - 2')

plotIndiv(final.splsda, comp = c(1,3), # plot samples from final model, c(1,2)
          group = diag, ind.names = FALSE, # colour by class label
          ellipse = TRUE, legend = TRUE, # include 95% confidence ellipse
          title = 'sPLS-DA on UC, comp 1 - 3')
```
```{r}
# set the styling of the legend to be homogeneous with previous plots
legend=list(legend = levels(diag), # set of classes
            col = unique(color.mixo(diag)), # set of colours
            title = "Tumour Type", # legend title
            cex = 0.7) # legend size
# Need to open X11() and expand window
# generate the CIM, using the legend and colouring rows by each sample's class
cim <- cim(final.splsda, row.sideColors = color.mixo(diag), 
           legend = legend)
```

```{r, echo=FALSE}
# form new perf() object which utilises the final model
perf.splsda.metabo <- perf(final.splsda, 
                          folds = 5, nrepeat = 10, # use repeated cross-validation
                          validation = "Mfold", dist = "max.dist",  # use max.dist measure
                          progressBar = FALSE)

# plot the stability of each feature for the first component, 'h' type refers to histogram
plot(perf.splsda.metabo$features$stable[[1]], type = 'h', 
     ylab = 'Stability', 
     xlab = 'Features', 
     main = '(a) Comp 1', las = 2)

plot(perf.splsda.metabo$features$stable[[2]], type = 'h', 
     ylab = 'Stability', 
     xlab = 'Features', 
     main = '(b) Comp 2', las = 2)

plot(perf.splsda.metabo$features$stable[[3]], type = 'h', 
     ylab = 'Stability', 
     xlab = 'Features', 
     main = '(c) Comp 3', las = 2)
```
```{r}
#var.name.short <- substr(srbct$gene.name[, 2], 1, 10) # form simplified gene names

plotVar(final.splsda, comp = c(1,2), var.names = list(diag), cex = 3) # generate correlation circle plot
```
```{r}
plotVar(final.splsda, comp = c(1,3), var.names = list(diag), cex = 3) # generate correlation circle plot
```
```{r}
library(rgl)
# undergo the pca method
scale.pca.metab <- pca(metab, ncomp = 3, center = TRUE, scale = TRUE) 

plotVar(scale.pca.metab, style = '3d') # plot a basic 3D correlation sphere plot
```

------------------------------------------------------------------------

### Training the model

```{r, include=FALSE}
# train the model
train.splsda.metabo <- splsda(metab.train, diag.train, ncomp = optimal.ncomp, keepX = optimal.keepX) 
```

```{r, include=FALSE}
# use the model on the Xtest set
predict.splsda.metabo <- predict(train.splsda.metabo, metab.test, 
  dist = "mahalanobis.dist")
```

#### Confusion matrix

```{r, include=FALSE}
# evaluate the prediction accuracy for the first two components
predict.comp <- predict.splsda.metabo$class$mahalanobis.dist[,3]
cmatrix <- table(factor(predict.comp, levels = levels(diag)), diag.test)
```


```{r}
knitr::kable(cmatrix, format = "simple", caption = "Confusion matrix")
```
```{r, echo=FALSE}
library(caret)
confusionMatrix(factor(predict.comp, levels = levels(diag)), diag.test, positive = "UC")
```
### Performance plots

```{r}
library(ConfusionTableR)
ConfusionTableR::binary_visualiseR(train_labels = factor(predict.comp, levels = levels(diag)),
                                   truth_labels= diag.test,
                                   class_label1 = "nonIBD", 
                                   class_label2 = "UC",
                                   quadrant_col1 = "#175E54", 
                                   quadrant_col2 = "#8C1515", 
                                   custom_title = "Confusion Matrix", 
                                   text_col= "black"
                                   )
```


```{r, echo=FALSE}
auc.splsda = auroc(final.splsda, roc.comp = 1, print = FALSE) # AUROC for the first component
auc.splsda = auroc(final.splsda, roc.comp = 3, print = FALSE) # AUROC for the three components
```

***

## Odds ratio for possible confounding variables
```{r}
# Primary variables of interest: consent_age, sex, race

library(MASS)
has_UC <- ifelse(diag == 'UC', 1, 0)
model= glm(formula = has_UC ~ samples$sex, family=binomial)
exp(cbind("Odds ratio" = coef(model), confint.default(model, level = 0.95)))

model= glm(formula = has_UC ~ samples$consent_age, family=binomial)
exp(cbind("Odds ratio" = coef(model), confint.default(model, level = 0.95)))

model= glm(formula = has_UC ~ samples$race, family=binomial)
exp(cbind("Odds ratio" = coef(model), confint.default(model, level = 0.95)))
```

